---
title: "Reproducible Research: Analysis of Storm data"
author: "Daniel Cevallos"
date: "12/05/2023"
output: 
  html_document:
    keep_md: true
---

Introduction Impact of Weather Events in the US
======================================================


Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

Synopsis
========

The analysis on the storm event database revealed that tornadoes are the most
dangerous weather event to the population health. While as the economic impact of weather events was
Flash floods and thunderstorm winds caused billions of dollars
in property damages. The largest crop damage is caused by
drought. All these things between years 1950 and 2011


Data Processing
===============

The analysis was performed on
[Storm Events Database](http://www.ncdc.noaa.gov/stormevents/ftp.jsp), provided by
[National Climatic Data Center](http://www.ncdc.noaa.gov/). The data is from a comma-separated-value file available
[here](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2).
There is also some documentation of the data available
[here](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf).

Initialize with the read of the data
```{r cache=TRUE}
stormData <- read.csv(bzfile("./data/repdata_data_StormData.csv.bz2"))
```

The Data Clean process involves case and punctuation changes. Verifying the number of events

```{r}
# number of unique event types
length(unique(stormData$EVTYPE))
stormData$EVTYPE <- toupper(stormData$EVTYPE)
stormData$EVTYPE <- gsub("[[:blank:][:punct:]+]", " ", stormData$EVTYPE)
# number of unique event types verified
length(unique(stormData$EVTYPE))
```

Dangerous Events with respect to Population Health
================================================

To find the event types that are most harmful to population health, the total number
of events are aggregated by the event type. After separate by fatalities and injuries. 

```{r}
library(plyr)
totalEvnt <- ddply(stormData, .(EVTYPE), summarize,
                    fatalities = sum(FATALITIES),
                    injuries = sum(INJURIES))

# Find events that caused most death and injury
fatality <- head(totalEvnt[order(totalEvnt$fatalities, decreasing = T), ], 10)
injury<- head(totalEvnt[order(totalEvnt$injuries, decreasing = T), ], 10)

```

Top 10 events that caused largest number of fatalities

```{r}
fatality[, c("EVTYPE", "fatalities")]
```

Top 10 events that caused most number of injuries

```{r}
injury[, c("EVTYPE", "injuries")]
```

Economic Effects of Weather Events
==================================

To analyze the impact on the economy, available property
damage and crop damage reportings/estimates were used.

In the raw data, the property damage is represented with two fields, a number
`PROPDMG` in USD and the exponent `PROPDMGEXP`. Similarly, the crop damage
is represented using two fields, `CROPDMG` and `CROPDMGEXP`. The first step in the
analysis is to calculate the property and crop damage for each event.

```{r}
transformUnits <- function(e) {
    # h -> hundred, k -> thousand, m -> million, b -> billion
    if (e %in% c('h', 'H'))
        return(2)
    else if (e %in% c('k', 'K'))
        return(3)
    else if (e %in% c('m', 'M'))
        return(6)
    else if (e %in% c('b', 'B'))
        return(9)
    else if (!is.na(as.numeric(e))) 
        return(as.numeric(e))
    else if (e %in% c('', '-', '?', '+'))
        return(0)
    else {
        stop("Invalid value.")
    }
}
```

```{r cache=TRUE}
prop_dmg_exp <- suppressWarnings(sapply(stormData$PROPDMGEXP, FUN=transformUnits))
stormData$prop_dmg <- stormData$PROPDMG * (10 ** prop_dmg_exp)
crop_dmg_exp <- suppressWarnings(sapply(stormData$CROPDMGEXP, FUN=transformUnits))
stormData$crop_dmg <- stormData$CROPDMG * (10 ** crop_dmg_exp)

```


The Economic losses by Event. 
```{r}
library(plyr)
losses <- ddply(stormData, .(EVTYPE), summarize,
                   prop_dmg = sum(prop_dmg),
                   crop_dmg = sum(crop_dmg))

# filter out events that caused no economic loss
losses <- losses[(losses$prop_dmg > 0 | losses$crop_dmg > 0), ]
prop_dmg_evnt <- head(losses[order(losses$prop_dmg, decreasing = T), ], 10)
crop_dmg_evnt <- head(losses[order(losses$crop_dmg, decreasing = T), ], 10)
```

Top 10 events that caused most property damage USD

```{r}
prop_dmg_evnt[, c("EVTYPE", "prop_dmg")]
```

Top 10 events that caused most crop damage

```{r}
crop_dmg_evnt[, c("EVTYPE", "crop_dmg")]
```

Results
=======

Health impact of weather events
-------------------------------

The following are the most dangerous weather event types.

```{r}
library(ggplot2)
library(gridExtra)
#fatalities events
p1 <- ggplot(data=fatality,
             aes(x=reorder(EVTYPE, fatalities), y=fatalities, fill=fatalities)) +
    geom_bar(stat="identity") +
    coord_flip() +
    ylab("Fatalities") +
    xlab("Event") +
    theme(legend.position="none")
#injuries events
p2 <- ggplot(data=injury,
             aes(x=reorder(EVTYPE, injuries), y=injuries, fill=injuries)) +
    geom_bar(stat="identity") +
    coord_flip() + 
    ylab("Injuries") +
    xlab("Event") +
    theme(legend.position="none")

grid.arrange(p1, p2, top="Most deadly weather events US between (1950-2011)")
```

As see, the `TORNADO` event have the most number of deaths and injuries. There 
are more than 5000 deaths and more than 10000 injuries between 1950 and 2011
in US. 
The next event types that are most dangerous with respect to population health
are `excessive heat` and `flash floods`.

Economic impact of weather events
---------------------------------

The following are the most severe weather event types


```{r}
library(ggplot2)
library(gridExtra)

p1 <- ggplot(data=prop_dmg_evnt,
             aes(x=reorder(EVTYPE, prop_dmg), y=log10(prop_dmg), fill=prop_dmg )) +
    geom_bar(stat="identity") +
    coord_flip() +
    xlab("Event") +
    ylab("Property Damage USD (log)") +
    theme(legend.position="none")

p2 <- ggplot(data=crop_dmg_evnt,
             aes(x=reorder(EVTYPE, crop_dmg), y=crop_dmg, fill=crop_dmg)) +
    geom_bar(stat="identity") +
    coord_flip() + 
    xlab("Event") +
    ylab("Crop Damage USD") + 
    theme(legend.position="none")

grid.arrange(p1, p2, top="Weather costs US between (1950-2011)")
```

Property damages are given in logarithmic scale.
The `FLASH FLOODS` and `THUNDERSTORM WINDS` cost the largest
property damages. Due to
untidy nature of the available data, type flood and flash flood are
separate values and should be merged for more accurate data-driven conclusions.

The most severe weather event in terms of crop damage is the `DROUGHT`. The drought has caused more than 10 billion USD damage followed by severe crop-damage-causing event types are floods and hails.
